---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(knitr.kable.NA = '.')
```

# tabshiftr <a href='https://ehrmanns.github.io/tabshiftr/'><img src='man/figures/logo.svg' align="right" height="200" /></a>


## Overview

Data are stored in many different ways in tables or spreadsheets because no strict semantic or topographic standards for the organisation of tables are commonly accepted.
In the R environment the *tidy* paradigm is a first step towards interoperability of data, in that it requires a certain arrangement of tables, where variables are recorded in columns and observations in rows (see https://tidyr.tidyverse.org/).
Tables can be tidied (i.e., brought into a tidy arrangement) via packages such as `tidyr`, however, all functions that deal with reshaping tables to date require data that are already organised into topologically coherent, rectangular tables.
This is often violated in practice, for example when data are scraped off of the internet.

`tabshiftr` fills this gap in the toolchain for reproducible data management via `schema` descriptions and a `reorganise()` function that is largely based on `tidyr`.
We call data that are not topologically coherent *disorganised (messy) data* and they are characterised not only by \"being messy\" (as in opposed to tidy), but also by being arranged so that not all variables are part of the same \"chunk\" of data or even that some variables are only available by implication.

## Installation

1) Install the latest development version from github:

```{r, eval=FALSE}
devtools::install_github("EhrmannS/tabshiftr")
```

2) The [vignette](https://ehrmanns.github.io/tabshiftr/articles/tabshiftr.html) provides an instruction on how to set up schema descriptions and shows a wide selection of table arrangements and which schemas apply to them.


## Examples

A disorganised table may look like the following table:

```{r}
library(readr)
library(tabshiftr)
library(knitr)

input <- read_csv(
  file = paste0(system.file("test_datasets", package = "tabshiftr"), "/table13.csv"),
  col_names = FALSE, col_types = cols(.default = "c"))
kable(input)
```

This table contains two chunks of data at the rows 1 to 6 and 8 to 13 and they are further organised into three similar groups (so-called *clusters*).
The clusters are somewhat systematic in that there are three variables in common, `commodities`, `harvested` and `production`.
The variable `year` has values but no explicit name and is not available for each cluster, and the variable `territories` is not even stored in a column, but is implicit and noted in the second row of each cluster.

In `tabshiftr` we distinguish between *identifying variables*, variables that describe an observational unit, and *measured variables*, variables that provide some value for that observational unit.
In the example,  `territories`, `year` and `commodities` are identifying variables and `harvested` and `production` are measured variables, they give the harvested area and the production of the given commodity at a given year in a given territory.

The approach of `tabshiftr` is based on capturing the arrangement of tables in so-called  [schema](https://en.wikipedia.org/wiki/Database_schema) description.
Typically there is an input and an output schema, describing the arrangement of the input and output tables, respectively.
The advantage of this procedure is that input and output tables exist explicitly and the schema maps the transformation of the data.
As we want to end up with tidy tables the output schema is pre-determined by a tidy table of the included variables, but the input schema description needs to be provided by the user.

For the input schema we have to describe, according to a [hierarchical set of rules](https://ehrmanns.github.io/tabshiftr/articles/tabshiftr.html#setting-up-schema-descriptions), in which rows and columns of a table which information can be found (see the example below).
First, we need to describe whether the table is organised in several clusters, and if so where they are located.
In the example, the \"origin\" of the first cluster is at the first row from the top and the first cell from left (`1|1`), the second at `1|8` and the third cluster originates at `4|8`.
All clusters are 3 cells wide and 6 cells high.
Moreover, we already recognised that each cluster represents one teritorial unit, so we record the variable `territories` as cluster ID.
A table should have some sort of header, i.e., one or more rows that describes which information a column contains and we need to register this header, in the example this is the first row of each cluster.
When there are several clusters that are identical in arrangement (but not in values), it makes sense to describe them in terms of `rel`ative values, i.e., starting to count from the origin of the cluster instead of the spreadsheet.

Input tables may contain many more data/variables than what we are interested in, but the schema description contains only those variables we want to have in the output table.
For each variable we need to define at least the `type` and either `col`umn(s), or column(s) and `row`(s).
Some variables are `dist`inct from the cluster outline, because they occur perhaps only once in the table, or are organised in a non-systematic way, such as the variable `year` in the example.
Some variables are either absent from the table, or are available only implicitly, for example if several spreadsheets or files contain information for one level of an identifying variable, such as per commodity.
This is not the case in the example, but if it were so, this variable could be specified by providing the `value` of the level in the variable.
For measured variables we also need to provide the target unit and a transformation factor to that unit, the latter of which will be multiplied with the values of the measured variable.
This results in the following schema description for our example:

```{r}
library(tabshiftr)
schema <- makeSchema(schema = list(
  clusters =
    list(row = c(1, 8, 8), col = c(1, 1, 4), width = 3, height = 6, id = "territories"),
  header = list(row = 1, rel = TRUE),
  variables =
    list(territories =
           list(type = "id", row = 1, col = 1, rel = TRUE),
         year =
           list(type = "id", row = c(3:6), col = 4, dist = TRUE),
         commodities =
           list(type = "id", col = 1, rel = TRUE),
         harvested =
           list(type = "measured", unit = "ha", factor = 1, col = 2, rel = TRUE),
         production =
           list(type = "measured", unit = "t", factor = 1, col = 3, rel = TRUE))
))
```

Some further fields that have not been mentioned here are:

* `split`, to seperate an identifying variable via a regular expression from some column
* `key` and `value`, when the names of measured variables are provided in a column while the respective values are in a seperate column. Here, `key` is the column that contains the names and `value` is the level of the measured variable that should be \"pulled\" out of `key`.

```{r}
schema
```

Finally, the input table is reorganised simply by calling `reorganise()`.

```{r}
output <- reorganise(input = input, schema = schema)
kable(output)
```

# To implement

